---
title: "gg_animate"
author: "Nick Koleits"
date: "08/02/2021"
output: html_document
---

# Iteration and animation: Loops, GIFs, and videos

This script servers as an introduction for creating animated GIFs and videos from charts made in R with ggplot2, using the gganimate package. See <https://paldhous.github.io/ucb/2018/dataviz/week14.html> for full article, along with original code.

When running this code in R Studio, when you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. 


## Load the required packages to the R Session. 
If not already installed, do so using **install.packages("package_name")**.

```{r, echo=FALSE}
# load required packages
library(readr)
library(ggplot2)
library(gganimate)
library(scales)
library(dplyr)
library(transformr)
library(viridis)
library(ggdark)
```

First, read the nations data into the R session. This data contains the GPD per capita and life expectancy (among other variables) which is used for creating both static and animated charts. 

```{r, echo=FALSE}
# load data
nations <- read_csv("data/nations.csv")
head(nations)
```
## GPD Life Expectance plot

The code below creates a static plot of GPD vs Life Expectancy for all nation for 2016. 

```{r gpd_life_expectancy_st, echo=FALSE}
# filter for 2016 data only
nations2016 <- nations %>%
  filter(year == 2016)

# make bubble chart
ggplot(nations2016, aes(x = gdp_percap, y = life_expect)) +
  xlab("GDP per capita ($)") +
  ylab("Life expectancy") +
  theme_minimal(base_size = 12) +
  geom_point(aes(size = population, color = region), alpha = 0.7) +
  scale_size_area(guide = FALSE, max_size = 15) +
  scale_x_continuous(labels = dollar) +
  stat_smooth(formula = y ~ log10(x), se = FALSE, size = 0.5, color = "black", linetype="dotted") +
  scale_color_brewer(name = "", palette = "Set2") +
  theme(legend.position=c(0.8,0.4))
```

The code below will create an animated version of the chart above showing the changes from 1990 to 2016. Running this code will create an R plot called  nations_plot. If running in R Studio, it will appear in the Viewer pane. 
```{r gpd_life_expectancy_an, echo=FALSE}
# animate entire time series with gganimate
nations_plot <- ggplot(nations, aes(x = gdp_percap, y = life_expect)) +
  xlab("GDP per capita") +
  ylab("Life expectancy at birth") +
  theme_minimal(base_size = 12, base_family = "Georgia") +
  geom_point(aes(size = population, color = region), alpha = 0.7) +
  scale_size_area(guide = FALSE, max_size = 15) +
  scale_x_continuous(labels = dollar) +
  stat_smooth(formula = y ~ log10(x), se = FALSE, size = 0.5, color = "black", linetype="dotted") +
  scale_color_brewer(name = "", palette = "Set2") +
  theme(legend.position=c(0.8,0.4)) +
  
  # gganimate code
  ggtitle("{frame_time}") +
  transition_time(year) +
  ease_aes("linear") +
  enter_fade() +
  exit_fade()

# code below displays the animated chart in the viewer pane
animate(nations_plot)
```

##How the gganimate code works

* **transition_time** animates the data by year variable, generating a frame (or static version) of the chart for each year. 
* **ggtitle("{frame_time}")** adds the year title to each frame.
* **ease_aes** controls how the animation progresses. If animating over a time series, always use the option "linear" to ensure a constant speed.
* **enter_fade** and **exit_fade* functions control the behavior where a data point appears or disappears from the animation.

##Save as a GIF and a video

Below saves the animation as a GIF.
```{r, echo=FALSE}
# save as a GIF
animate(nations_plot, fps = 10, width = 750, height = 450)
anim_save("animations/nations.gif")
```

##Make a cumulative animation of historical global average temperature

The code below demonstrates a scatter plot style of chart and animation. Using a new dataset, **warming**, we will plot the difference from the mean temperature for 1900-2000, from 1880 to 2017. 

```{r warming_st, echo=FALSE, fig.cap="Degree deviation from mean temperature (1900-2000), by year"}
# load data
warming <- read_csv("data/warming.csv")

# inspect data
head(warming)

# draw chart
w_plot <- ggplot(warming, aes(x = year, y = value)) +
  geom_line(colour="grey") +
  geom_point(shape = 21, colour = "black", aes(fill = value), size=5, stroke=1) +
  scale_x_continuous(limits = c(1880,2017)) +
  scale_y_continuous(limits = c(-0.5,1)) +
  scale_fill_distiller(palette = "YlGnBu", limits = c(-0.5,1), guide = FALSE) +
  xlab("Year") +
  ylab("Difference from average temp 1900-2000 (ºC)") +
  theme_minimal()

w_plot
```

The file warming.csv contains the variables year and value. Value is the global annual average temperature, compared to the average global temperature from 1900-2000.

The code below creates an animated point version of the chart above, viewable in the R Studio Viewer pane and saved as a gif. 
```{r warming_an, echo=FALSE}
# draw chart
warming_points <- ggplot(warming, aes(x = year, y = value)) +
  geom_point(shape = 21, colour = "black", aes(fill = value), size=5, stroke=1) +
  scale_x_continuous(limits = c(1880,2017)) +
  scale_y_continuous(limits = c(-0.5,1)) +
  scale_fill_distiller(palette = "YlGnBu", limits = c(-1,1), guide = FALSE) +
  xlab("Year") +
  ylab("Difference from average temp 1900-2000 (ºC)") +
  theme_minimal(base_size = 16) +
  # gganimate code
  transition_time(year) +
  shadow_mark()

# save as a GIF
animate(warming_points, fps = 10, width = 750, height = 450)
anim_save("animations/warming_points.gif")
```
Since the static version of the chart above used both geom_point and geom_line layers, we need to create individual plots for each year and convert to a gif to create the animated version of the plot. The code below loops over all years  **years <- c(1880:2017)** in the data and creates a single plot for each, saving it to file.

```{r, echo=FALSE}
# make a list of years, from 1880 to 2017
years <- c(1880:2017)

# loop to make a chart for each year
for (y in years) {
  tmp <- warming %>%
    filter(year <= y)
  chart <- ggplot(warming, aes(x = year, y = value)) +
  geom_line(colour="grey") +
  geom_point(shape = 21, colour = "black", aes(fill = value), size=5, stroke=1) +
  scale_x_continuous(limits = c(1880,2017)) +
  scale_y_continuous(limits = c(-0.5,1)) +
  scale_fill_distiller(palette = "YlGnBu", limits = c(-0.5,1), guide = FALSE) +
  xlab("Year") +
  ylab("Difference from average temp 1900-2000 (ºC)") +
  theme_minimal()
  ggsave(file = paste0("charts/",y,".jpg"), plot = chart, width = 8, height = 4.5, units = "in", dpi = 300)
  print(paste0("processing: ",y))
}

```
Another useful type of animation is one that switches between two states to show the difference between two categories, periods of any other variables of comparison. The chart below switches between a simulation of human effects on global average temperature, and natural ones

The **simulations** data below shows a simulation based on climate models of how the global average temperature would have changed under the influence of natural events, such as variation in radiation from the Sun and the cooling effect of soot from volcanoes, compared to human influences, mostly emissions of carbon dioxide and other greenhouse gases.

First, read in and inspect the data.

```{r, echo=FALSE}
# load data
simulations <- read_csv("data/simulations.csv")

# inspect data
head(simulations)
```
##How the gganimate code works

**transition_state.** This switches between different filtered states of the data, here defined by the variable type. transition_length is the length of the transition in seconds, and state_length is the pause at each state, again in seconds.

**ease_aes** With a state transition animation, using options that vary the speed of the transition, with a slower start and finish than the middle section, give a more visually pleaseing effect. Try "cubic-in-out" or "sine_in_out"

Using **"{closest_state}**" in the ggtitle function displays the appropriate value for the variable used to define the states, here type.


This code will load the data and make the animation:

```{r simulation_an, echo=FALSE}
# draw chart
simulations_plot <- ggplot(simulations, aes(x=year, y=value, color = value)) +
  geom_line(size = 1.5) +
  scale_y_continuous(limits = c(-0.6,0.75)) +
  scale_colour_distiller(palette = "RdYlBu", limits = c(-0.4,0.7), guide = FALSE) +
  ylab("Difference from 1900-2000 average (ºC)") +
  xlab("Year") +
  dark_theme_dark(base_size = 16) +
  #gganimate code
  ggtitle("{closest_state}") +
  transition_states(
    type,
    transition_length = 0.5,
    state_length = 2
  ) +
  ease_aes("sine-in-out")

# save as a GIF
animate(simulations_plot, fps = 10, width = 750, height = 450)
anim_save("animations/simulations.gif")
```

